Combining predicate transformer semantics for effects: a case study in parsing regular languages

### 2020-08-25 msfp2020/msfp2020.tex
Update slides

\titlepage

\begin{frame}{The history}

\emph{Predicate transformer semantics} for imperative programs have been around for 45 years.
A statement becomes a map from postcondition (a predicate on the state space) to the weakest precondition that ensures the postcondition will be satisfied.

\emph{Parser combinators} have been around in functional programming for 35 years.
Parsing becomes an effectful (monadic) operation, nondeterministically modifying state (the unparsed part of the string).

\emph{Algebraic effects} have recently gained traction in the programming language community.

\begin{frame}[fragile]{Algebraic effects}

@@

\begin{minted}{Agda}
data Free (C : Set) (R : C -> Set) : Set -> Set where
	Pure : a -> Free C R a
	Op : (c : C) -> (k : R c -> Free C R a) -> Free C R a

data CNondet where
	Fail : CNondet
	Choice : CNondet
RNondet : CNondet -> Set
RNondet Fail = ⊥
RNondet Choice = Bool

@@

\begin{frame}[fragile]{Semantics for algebraic effects}
\emph{Handlers} give semantics for the \Free monad naturally as a fold:

\begin{minted}{Agda}
handleList : Nondet a -> List a
handleList (Pure x) = [x]
handleList (Op Fail k) = []
handleList (Op Choice k) = k True ++ k False
\end{minted}

The generic fold that computes a predicate of type \Set:

\begin{minted}{Agda}
[[_]] : Free C R a -> ((c : C) -> (R c -> Set))
      -> (a -> Set) -> Set
[[ Pure x ]] alg P = P x
[[ Op c k ]] alg P = alg c (λ x -> [[ k x ]] alg P)
\end{minted}
\end{frame}

@@

\begin{frame}[fragile]{Parsing regular expressions}

Consider the following syntax of regular expressions and their parse trees:
\begin{minted}{Agda}
data Regex : Set where
    Empty : Regex
    Epsilon : Regex
    Singleton : Char → Regex
    _ | _ : Regex → Regex → Regex
    _ · _ : Regex → Regex → Regex
    _ * : Regex → Regex

Tree : Regex -> Set
Tree Empty = ⊥
Tree Epsilon = ⊤
Tree (Singleton _) = Char
Tree (l | r) = Either (Tree l) (Tree r)
Tree (l · r) = Pair (Tree l) (Tree r)
Tree (r *) = List (Tree r)
\end{minted}

Let's write a parser returning these \Agda{Tree}s.
\end{frame}

\begin{frame}[fragile]{Parsing regular expressions}
A nondeterministic \Regex matcher is straightforward to write:
\begin{minted}{Agda}
match : (r : Regex) -> String -> Nondet (Tree r)
match Empty xs = Op Fail λ()
match Epsilon Nil = Pure tt
match Epsilon (_ :: _) = Op Fail λ()
match (Singleton c) xs =
    if xs = [c] then Pure c else Op Fail λ()
match (l | r) xs = Op Choice (λ b,
    if b then Inl <$> match l xs else Inr <$> match r xs)
match (l · r) xs = do
    (ys, zs) <- allSplits xs
    (,) <$> match l ys <*> match r zs
\end{minted}

\pause
... until we reach the last case:
\begin{minted}{Agda}
match (r *) xs = match (r · (r *)) xs
    -- Error: does not terminate
\end{minted}

@@

A predicate transformer \Agda{pt1} \emph{is refined by} \Agda{pt2} if \Agda{pt₂} satisfies more postconditions than \Agda{pt₁}:
\vspace{-\baselineskip}
\begin{minted}{Agda}
_⊑_ : (pt1 pt2 : (a -> Set) -> Set) -> Set
pt1 ⊑ pt2 = ∀ P -> pt1 P -> pt2 P
\end{minted}

The relation \Agda{S ⊑ T} expresses that the program \Agda{T} is ``better'' than \Agda{S}:
\Agda{S} can be replaced with \Agda{T} everywhere, and all postconditions will still hold.

@@

By assigning a predicate transformer to specifications, we can also relate specifications and programs:
\vspace{-\baselineskip}
\begin{minted}{Agda}
[[_,_]] : (pre : Set) (post : a -> Set) -> (a -> Set) -> Set
[[ pre , post ]] P = pre ∧ ∀ x, post x -> P x
\end{minted}

This is the `weakest precondition' necessary so that the desired post- condition \Agda{P} holds:
\Agda{pre} should hold and any result satisfying \Agda{post} should imply the postcondition \Agda{P}.

@@

\begin{frame}[fragile]{Adding effects}
The problem with \match is that implementing the Kleene star requires the effect of \emph{general recursion}.

We can add more effects to the free monad by choosing the command and response types from a list of \emph{effect signatures}:
\begin{minted}{Agda}
data Free (es : List Sig) : Set -> Set where
	Pure : a -> Free es a
	Op : (i : mkSig C R ∈ es) (c : C) -> (k : R c -> Free C R a) -> Free C R a
\end{minted}

We will add two new effects along with nondeterminism: general recursion and parsing.
\end{frame}

@@

The semantics of \Agda{Parser} are stateful: returning the next character of the input string requires keeping track of the remaining characters.
This state can be found in the \Agda{ptParser} semantics as an extra argument to the predicates.
\begin{minted}{Agda}
ptParser : ⊤ -> (Maybe Char -> String -> Set) -> String -> Set
ptParser _ P Nil = P Nothing Nil
ptParser _ P (x :: xs) = P (Just x) xs
\end{minted}

@@

\begin{frame}[fragile]{Defining a derivative-based matcher}
\begin{minted}{Agda}
dmatch : (Forall(es)) ⦃ iP : Parser ∈ es ⦄ ⦃ iND : Nondet ∈ es ⦄  -> (RecArr Regex es tree)
dmatch r = symbol >>= maybe
    (λ x -> integralTree r <$> call (hiddenInstance(∈Head)) (d r /d x))
    (if p <- matchEpsilon r then Pure (Sigma.fst p) else (hiddenConst(fail)))
\end{minted}
\end{frame}

\begin{frame}[fragile]{Termination checking}
\begin{minted}{Agda}
terminates-in : (Forall(I O es a)) (pts : PTs es) (f : (RecArr I es O)) (S : Free (Rec I O :: es) a) → Nat → Set
terminates-in pts f (Pure x)            n         = ⊤
terminates-in pts f (Op ∈Head c k)      Zero      = ⊥
terminates-in pts f (Op ∈Head c k)      (Succ n)  = terminates-in pts f (f c >>= k) n
terminates-in pts f (Op (∈Tail i) c k)   n        =
    lookupPT pts i c (λ x → terminates-in pts f (k x) n)
\end{minted}
\end{frame}

\begin{frame}[fragile]{Refining \match with \Agda{dmatch}}
\begin{minted}{Agda}
dmatchSound : ∀ r xs -> (wpMatch (match (hiddenInstance(∈Head)) (r , xs))) ⊑ (wpMatch (dmatch' (hiddenInstance(∈Head)) (r , xs)))
\end{minted}
\end{frame}

>>>>

\titlepage
\begin{frame}[fragile]{Algebraic effects}

@@

\begin{minted}{Agda}
data Free (C : Set) (R : C -> Set) : Set -> Set where
	Pure : a -> Free C R a
	Op : (c : C) -> (k : R c -> Free C R a) -> Free C R a
\end{minted}
\end{frame}

\begin{frame}[fragile]{Example: Nondeterminism}

\Agda{Nondet} has two primitive operations:
\begin{itemize}
    \item \Agda{Choice} chooses between two values
    \item \Agda{Fail} goes to a failure state and stops execution
\end{itemize}
\begin{minted}{Agda}
data CNondet where
	Choice : CNondet
	Fail : CNondet

RNondet : CNondet -> Set
RNondet Choice = Bool
RNondet Fail = ⊥

@@

\begin{frame}[fragile]{Semantics for algebraic effects}
\emph{Handlers} give semantics for the \Free monad naturally as a fold:
\begin{minted}{Agda}
handleList : Nondet a -> List a
handleList (Pure x) = [x]
handleList (Op Choice k) = k True ++ k False
handleList (Op Fail k) = []
\end{minted}

\pause
The generic fold that computes a predicate of type \Set:
\begin{minted}{Agda}
[[_]] : Free C R a -> ((c : C) -> (R c -> Set))
      -> (a -> Set) -> Set
[[ Pure x ]] alg P = P x
[[ Op c k ]] alg P = alg c (λ x -> [[ k x ]] alg P)
\end{minted}
\end{frame}

@@

\begin{frame}[fragile]{Parsing regular expressions}
To illustrate these semantics, we wrote a parser.
The input is a regular expression and a \Agda{String}, and the output a parse tree.
\begin{minted}{Agda}
data Regex : Set where
    Empty : Regex
    Epsilon : Regex
    Singleton : Char → Regex
    _ | _ : Regex → Regex → Regex
    _ · _ : Regex → Regex → Regex
    _ * : Regex → Regex

Tree : Regex -> Set
Tree Empty         = ⊥
Tree Epsilon       = ⊤
Tree (Singleton _) = Char
Tree (l | r)       = Either (Tree l) (Tree r)
Tree (l · r)       = Pair (Tree l) (Tree r)
Tree (r *)         = List (Tree r)
\end{minted}
\end{frame}

\begin{frame}[fragile]{Parsing regular expressions}
\begin{minted}{Agda}
match : (r : Regex) -> String -> Nondet (Tree r)
match Empty         xs       = Op Fail λ()
match Epsilon       Nil      = Pure tt
match Epsilon       (_ :: _) = Op Fail λ()
match (Singleton c) xs       =
    if xs = [c] then Pure c else Op Fail λ()

match (l | r)       xs       = Op Choice (λ b ->
    if b then Inl <$> match l xs else Inr <$> match r xs)

match (l · r) xs = do
    (ys, zs) <- allSplits xs
    (,) <$> match l ys <*> match r zs
\end{minted}
Looks straightforward so far...

\pause
... until we reach the last case:
\vspace{-\baselineskip}
\begin{minted}{Agda}
match (r *) xs = match (r · (r *)) xs
    -- Error: does not terminate
\end{minted}

@@

A predicate transformer \Agda{pt1} \emph{is refined by} \Agda{pt2} if \Agda{pt2} satisfies more postconditions than \Agda{pt1}:
\vspace{-\baselineskip}
\begin{minted}{Agda}
_⊑_ : (pt1 pt2 : (a -> Set) -> Set) -> Set
pt1 ⊑ pt2 = ∀ P -> pt1 P -> pt2 P
\end{minted}

\Agda{S ⊑ T} expresses that \Agda{T} is ``better'' than \Agda{S}:
\Agda{S} can be replaced with \Agda{T} everywhere, and all postconditions will still hold.

@@

Predicate transformers are a semantic domain where programs and specifications can be related.
\vspace{-\baselineskip}
\begin{minted}{Agda}
[[_,_]] : (pre : Set) (post : a -> Set) -> (a -> Set) -> Set
[[ pre , post ]] P = pre ∧ ∀ x, post x -> P x
\end{minted}

@@

\begin{frame}[fragile]{Adding effects}
The problem with \match is that implementing the Kleene star also requires the effect of \emph{general recursion}.

We can add more effects to the free monad by choosing the command and response types from a list of \emph{effect signatures}:
\begin{minted}{Agda}
data Free (es : List Sig) : Set -> Set where
    Pure : a -> Free es a
    Op : (i : mkSig C R ∈ es) (c : C)
         (k : R c -> Free C R a) -> Free C R a
\end{minted}

We will add two new effects: general recursion and parsing.
\end{frame}

@@

\Agda{Parser} has stateful semantics: to return the next character, we need to keep track of the remaining characters.
The state is the extra \Agda{String} arguments in \Agda{ptParser}.
\begin{minted}{Agda}
ptParser : ⊤ -> (Maybe Char -> String -> Set) -> String -> Set
ptParser _ P Nil = P Nothing Nil
ptParser _ P (x :: xs) = P (Just x) xs
\end{minted}

@@

\begin{frame}[fragile]{Extending \match}

Now we can finish the definition and prove soundness unconditionally:
\begin{minted}{Agda}
match (r *) = Op iRec (r · (r *))

matchSound : (r : Regex) (xs : String) ->
    [[ ⊤ , post r xs ]] ⊑ [[ match r xs ]]
\end{minted}

\pause
\match still does not terminate if \Agda{r} matches the empty string, our result is only \emph{partial correctness}.\\
\Agda{ptRec} computes the WLP: all recursive calls immediately return.
\end{frame}

\begin{frame}[fragile]{Defining a derivative-based matcher}
To guarantee termination, use recursion on \Agda{xs} rather than \Agda{r}.\\
The \emph{Brzozowski derivative} \Agda{d r /d x} matches \Agda{xs} iff \Agda{r} matches \Agda{x :: xs};\\ \Agda{integralTree r : tree (d r /d x) -> tree r} ``integrates'' parse trees.

\begin{minted}{Agda}
dmatch : (r : Regex) -> Free es (tree r)
dmatch r = symbol >>= maybe
    (λ x -> Op iRec (d r /d x) (integralTree r))
    (if p <- matchEpsilon r
     then Pure (Sigma.fst p)
     else Op iND Fail λ())
\end{minted}

\begin{minted}{Agda}
dmatchSound : ∀ r xs -> [[ match r xs ]] ⊑ [[ dmatch r xs ]]
\end{minted}
\end{frame}

\begin{frame}[fragile]{Termination checking}
\Agda{ptRec} gives weakest liberal precondition semantics.
For total correctness, we should check termination.

\vspace{\baselineskip}
\Agda{terminates-in f S n} holds iff \Agda{S} terminates after calling \Agda{f} at most \Agda{n} times.
\vspace{-\baselineskip}

\begin{minted}{Agda}
terminates-in : (f : (i : I) -> Free (Rec I O :: es) (O i))
    (S : Free (Rec I O :: es) a) → ℕ → Set
terminates-in f (Pure x)           n        = ⊤
terminates-in f (Op ∈Head c k)     Zero     = ⊥
terminates-in f (Op ∈Head c k)     (Succ n) =
    terminates-in pt f (f c >>= k) n
terminates-in f (Op (∈Tail i) c k) n        =
    pts i c (λ x → terminates-in f (k x) n)
\end{minted}
\end{frame}

\begin{frame}[fragile]{Total correctness}

Partial correctness of \Agda{dmatch} follows from the chain of refinements:
\begin{minted}{Agda}
[[ ⊤ , post r xs ]]
    ⊑ [[ match r xs ]]
    ⊑ [[ dmatch r xs ]]
    ⊑ [[ ⊤ , post r xs ]]
\end{minted}
together with a proof of termination:
\begin{minted}{Agda}
dmatchTerminates : (r : Regex) (xs : String) ->
    terminates-in dmatch (dmatch r xs) (length xs)
\end{minted}

\begin{frame}{Discussion}

In our paper, we illustrate how techniques from the refinement calculus can be used in functional programming.
A distinguishing characteristic of our approach is modularity:
we add new effects and semantics to the system as we need them.

Formally verified parsers have been developed before,
using specialized semantics to the domain of parsing.
The modularity of predicate transformers allow us to reason about effects uniformly.

Most existing approaches to recursion in parsers deal with termination syntactically.
Separation of syntax and semantics also cleanly separates partial and total correctness.
